---
title: "FMIS Geocoding Analysis"
author: "Finn Meffe"
date: "`r Sys.Date()`"
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-3em}
  - \pretitle{\begin{center}\large\bfseries}
  - \posttitle{\end{center}}
  - \preauthor{\begin{center}\normalsize}
  - \postauthor{\end{center}}
  - \predate{\begin{center}\normalsize}
  - \postdate{\end{center}}
output:
  pdf_document: 
    extra_dependencies: ["booktabs", "multirow", "makecell", "float"]
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
  encoding = encoding, 
  output_dir = "C:/Users/fm557/YLS Dropbox/Finn Meffe/FHWA cost data/Output/Finn", 
  output_file = paste0(format(Sys.Date(), "%Y-%m-%d"), " FMIS Geocoding Analysis.pdf"))
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 0, digits = 6)  
if(!require('pacman')) {
  install.packages('pacman')
}
pacman::p_load(ggplot2, dplyr, tidyverse, data.table, lubridate, ggpubr, skimr, scales, tidycensus, readxl, httr, jsonlite, sf, mapview, tigris, lehdr, spatstat, crsuggest, dbmss, stargazer, rayshader, lmtest, sandwich, fixest, modelsummary, usmap, haven, janitor, fredr, purrr, tidycensus, here, ggtext)

root_dir = "C:/Users/fm557/YLS Dropbox/Finn Meffe/FHWA cost data"
data_dir = file.path(root_dir, "Data", "Intermediate")
output_dir = file.path(root_dir, "Output", "Finn", "geocoding analysis")
code_dir = file.path(root_dir, "Code", "FMIS_finn", "04_figures_and_analysis", "geocoding_analysis")

knitr::opts_knit$set(root.dir = root_dir)

options(tinytex.latexmk.warning = FALSE)
```

```{r, eval=F, echo=F}
source(file.path(code_dir, "geocoding_load.R"), local=T) # takes awhile to parse the large .dta files, don't need to run unless changing the underlying joined project level data
```

```{r, warning=F, message=F, include=F}
source(file.path(code_dir, "geocoding_clean.R"), local=T)
```

```{r, include=F, eval=F}
# There seem to be slight discrepancies when comparing the total_cost_mills variable from the FMIS project data to the sum of costs from the FMIS receipts data (23 projects have a difference.)

test <- project_level_fmis_geo %>%
  inner_join(project_level_fmis, by = "project_id") %>% 
  filter(fp_ic == 1,
         completion_year >= 1950 & completion_year <= 2000,
         )

resid_test <- inner_join(
  plf_join %>% 
    select(project_id, total_cost_mills),
  test %>% 
    select(project_id, total_cost_mills),
  by = "project_id"
) %>% 
  mutate(diff = abs(total_cost_mills.x - total_cost_mills.y))

summary(resid_test)

non_zero_resids <- resid_test %>%
  filter(diff > 0) %>% 
  pull(project_id)
```

```{r map-style, include=F}
source(file.path(code_dir, "functions_aesthetic.R"), local=T)
```

# Summary {.unlisted .unnumbered}

This memo contains figures that explore potential geocodeability in FMIS projects. "Geocodeability" was assessed using Gemini title parsing to identify end points in project data (and the quality of these endpoints). Throughout these figures, I use "geocoded" as the binary outcome of interest, but it is important to note that this is just potential, rather than projects that have actual GIS coordinates. This title parsing was run on the complete sample of projects that were found in countries in which there was only one year of PR511 data ever. Our working definition of "geocodeability" here is if there are two identifiable endpoints and at least one has a high enough precision (we parameterize this at 3, defined [here](https://docs.google.com/document/d/1hBzvypsyAz6U1PnYGRJSiGCRTz1rKjRyHWh4ISOkCgw/edit?usp=sharing)). I find that:

- The distribution of geocoding rates (at the county level) is typically clumped at the extremes: there are many counties with no geocoding and a moderate amount with a high or full percentage.
- Counties that have high amounts of geocodeability (75\%+ of project spending geocoded) typically have *less* new construction, with receipts containing PE and ROW costs appearing more likely to be geocoded
- States and counties differ dramatically in geocodeability, and the final sample of counties that has both a single year of interstate construction and 75\%+ of spending geocoded is quite small, at n = 59.

\tableofcontents

\newpage
# Histograms

```{r hist-state, warning=F, include=F}
source(file.path(code_dir, "make_histograms.R"), local=T)
```

## State-level

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_proj_hist)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_spend_hist)
```

## County-level

```{r, echo=F, warning=F, fig.width = 6, fig.height = 4}
print(county_proj_hist)
```

```{r, echo=F, warning=F, fig.width = 6, fig.height = 4}
print(county_spend_hist)
```

```{r, include=F}
ggsave(file.path(output_dir, "hist_state_pct_proj_geocoded.png"),
       state_proj_hist, width = 8, height = 5)
ggsave(file.path(output_dir, "hist_state_pct_spending_geocoded.png"),
       state_spend_hist, width = 8, height = 5)
ggsave(file.path(output_dir, "hist_county_pct_proj_geocoded.png"),
       county_proj_hist, width = 8, height = 5)
ggsave(file.path(output_dir, "hist_county_pct_spending_geocoded.png"),
       county_spend_hist, width = 8, height = 5)
```

### By Decade

```{r, echo=F, warning=F, fig.width = 6, fig.height = 4}
print(county_proj_hist_decade)
```

```{r, echo=F, warning=F, fig.width = 6, fig.height = 4}
print(county_spend_hist_decade)
```

```{r, include=F}
ggsave(file.path(output_dir, "hist_county_pct_proj_geocoded_by_decade.png"),
       county_proj_hist_decade, width = 8, height = 5)
ggsave(file.path(output_dir, "hist_county_pct_spending_geocoded_by_decade.png"),
       county_spend_hist_decade, width = 8, height = 5)
```

\newpage
# Cost Breakdowns

## Cost Composition Breakdown by Geocoding Rate

```{r, warning=F, include=F}
source(file.path(code_dir, "make_cost_breakdowns.R"), local=T)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(cost_break_bar_only)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(cost_break_bar_any)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(cost_break_ungeocoded)
```

```{r, include=F}
ggsave(file.path(output_dir, "cost_type_breakdown_by_geocoded_all_counties.png"),
       cost_break_bar_only)
ggsave(file.path(output_dir, "cost_type_breakdown_by_geocoded_any_counties.png"),
       cost_break_bar_any)
```


## Geocoding Rate Breakdown by Cost Composition

### Share of Projects

```{r, echo=F, fig.width = 6, fig.height = 4}
print(geo_by_type_bar)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(ngeo_by_type_bar)
```

### Share of Projects (75+ Counties)

```{r, echo=F, fig.width = 6, fig.height = 4}
print(geo_by_type_bar_75any)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(ngeo_by_type_bar_75any)
```

### Share of spending

```{r, echo=F, fig.width = 6, fig.height = 4}
print(geo_spending_by_type_bar)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(ngeo_spending_by_type_bar)
```

### Share of spending (75+ Counties)

```{r, echo=F, fig.width = 6, fig.height = 4}
print(geo_spending_by_type_bar_75any)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(ngeo_spending_by_type_bar_75any)
```

```{r, include=F}
ggsave(file.path(output_dir, "geocoded_breakdown_by_cost_type.png"),
       geo_by_type_bar)
ggsave(file.path(output_dir, "not_geocoded_breakdown_by_cost_type.png"),
       ngeo_by_type_bar)
ggsave(file.path(output_dir, "geocoded_spending_breakdown_by_cost_type.png"),
       geo_spending_by_type_bar)
ggsave(file.path(output_dir, "not_geocoded_spending_breakdown_by_cost_type.png"),
       ngeo_spending_by_type_bar)
```

\newpage
# Maps

## State-level

```{r state-maps, include=F}
source(file.path(code_dir, "make_state_county_maps.R"), local=T)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_proj_map)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_spend_map)
```

### By Decade

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_proj_map_decade)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(state_spend_map_decade)
```

```{r, include=F}
ggsave(file.path(output_dir, "map_state_pct_recpt_geocoded.png"),
       state_proj_map, width = 8, height = 5)
ggsave(file.path(output_dir, "map_state_pct_spending_geocoded.png"),
       state_spend_map, width = 8, height = 5)
ggsave(file.path(output_dir, "map_state_pct_recpt_geocoded_by_decade.png"),
       state_proj_map_decade, width = 8, height = 5)
ggsave(file.path(output_dir, "map_state_pct_spending_geocoded_by_decade.png"),
       state_spend_map_decade, width = 8, height = 5)
```

## County-level

```{r, echo=F, fig.width = 6, fig.height = 4}
print(county_proj_map)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(county_spend_map)
```

```{r, echo=F, fig.width = 6, fig.height = 4}
print(county_spend_map_75_only)
```

```{r, include=F}
ggsave(file.path(output_dir, "map_county_pct_recpt_geocoded.png"),
       county_proj_map, width = 8, height = 5)
ggsave(file.path(output_dir, "map_county_pct_spending_geo.png"),
       county_spend_map, width = 8, height = 5)
ggsave(file.path(output_dir, "map_county_pct_spending_geo_75_only.png"),
       county_spend_map_75_only, width = 8, height = 5)
```

